---
title: "**The Bayesian Inferential Paradigm in Archaeology**"
author: 
  - ERIK OTÁROLA-CASTILLO*, MELISSA G. TORQUATO
  - Department of Anthropology, Purdue University, West Lafayette, Indiana, USA
  - CAITLIN E. BUCK*
  - School of Mathematics and Statistics, University of Sheffield, Sheffield, UK
date: "*R Markdown version last compiled on `r format(Sys.time(), '%A %B %d %Y, %X, %Z')`*"
indent: yes
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage[labelfont=bf,width=1\textwidth,justification=raggedright,singlelinecheck=false]{caption}
    - \captionsetup[figure]{labelfont=bf}
    - \usepackage[pagewise]{lineno}
output: 
  pdf_document: 
    latex_engine: xelatex
    keep_md: false
  word_document: default
  html_document: default
link-citations: yes
bibliography: cite.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(list(scipen=5, digits = 2))
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
########################################
# check for required packages          #
########################################
#setwd(dirname(rstudioapi:::getActiveDocumentContext()$path))
list.of.packages <- c("tidyverse","tidybayes", "rstan", "ggdist", "ggplot2", "cowplot", "knitr", "kableExtra", "tinytex")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
########################################
# load packages                        #
########################################
require(tidyverse)
require(tidybayes)
require(rstan)
require(ggdist)
require(ggplot2)
require(cowplot)
```

\vspace{4cm}
\noindent
**Manuscript accepted to the Handbook of Archaeological Sciences, 2nd ed. Forthcoming volume under contract (2022). *Edited by* M. Pollard, R.A. Armitage, and C.M. Makarewicz. Wiley.** 

\vspace{2.5cm}
\textbf{*\*Corresponding Authors}
\noindent
\newline
email:\href{mailto:eoc@purdue.edu}{\color{blue} \,eoc@purdue.edu}\newline 
email:\href{mailto:c.e.buck@sheffield.ac.uk}{\color{blue} \,c.e.buck@sheffield.ac.uk}\newline


\newpage
```{r child= "rmd_files/Introduction.Rmd"}

```
\newpage
```{r child= "rmd_files/A Simple Archaeological Example.Rmd"}

```
\newpage
```{r child= "rmd_files/What is Bayes' Theorem.Rmd"}

```
\newpage
```{r child= "rmd_files/Other Archaeologial Applications.Rmd"}

```
\newpage
```{r child= "rmd_files/Some Practicalities.Rmd"}

```
\newpage
```{r child= "rmd_files/Hopes for the Future.Rmd"}

```




